Incremental tactics

### 2012-05-11 doc/papers/LFMTP2012/paper.tex
paper (background)

>>>>

\paragraph{Motivational example}

A student is learning about System \sysname T, and wants to write a
program computing $(2+2)\times 3$. He (or she) knows that sometimes those
$\cst{rec}$ arguments are quite hard to get correct, but System
\sysname T is a typed language; he wants to take advantage of this to
build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
$$
For this, he sends the command $\infer{P_1}$ to the system. The
$\infer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call them $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\get\md$ from the derivation to the program it types, what needs really
to be sent to the system is:
\begin{align*}
  &\infer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\get{\md_4}}}
    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\infer{\get\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\get{\md_4}}$, $\md_6$ for the arguments
$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \infer{}(
  &\letb{add}{\get{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\get{\gsubst{\md_7}{\msubst H {\infer{\var{add}}}}})}
  {(\s{\get{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\infer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\infer{}(
\letb{add}{\get{\md_5}}
\get{({\gsubst{\md_9} {\msubst {H} {(\get{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\get{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\get{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

\paragraph{Discussion}

% scale to big programs
% memoization ds le cas HO
%   nécessité de fcts inverses?
% séparer generation des diffs et vérification
